---
title: "species richness comparison"
author: "Erica Christensen"
date: "4/5/2021"
output: html_document
---

```{r setup, include=FALSE}
#library(dplyr)
#library(ggplot2)
library(tidyverse)
library(broom)

# colors for figures
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# read data (written in species_richnes_difference.R)
richness_comparison = read.csv('richness_comparison_1928_2016.csv')

```

## Data

The question is how did species richness change between 1928 and 2016. I am using 74 quadrats that were no shrub or sparse shrub in 1928, and comparing species richness on the same quadrats in 2016. Of the 74, 22 quadrats were medium/high shrub encroached by 2016 (according to the remote sensed shrub map produced by Wenji) and 52 remained low-shrub. 

First of all, is the change in richness per quadrat between 1928 and 2016 significantly different from 0?

```{r cars, echo=F}
head(dplyr::select(richness_comparison, quadrat, richness_1928, richness_2016, sprich_change, shrubland))

# test whether the data are normally distributed
shapiro.test(richness_comparison$sprich_change)

# perform one-sample t-test: different from 0?
res <- t.test(richness_comparison$sprich_change, mu=0)
res

```

The test indicates that the difference is significantly different from zero, and on average each quadrat lost 1.05 species.

## Shrub encroached vs. not encroached
Plot the data by shrub-encroached or not-shrub-encroached:

```{r boxplor, echo=F}
# how many in each category
table(richness_comparison$shrubland)

# boxplot: richness comparing 1928 to 2016 and transitioned to shrub/ did not transition to shrub
box_richness_shrubtransition = ggplot(richness_comparison, aes(x=shrubland, y=sprich_change, group=shrubland, color=as.factor(shrubland))) +
  scale_color_manual(values=cbPalette[c(2,6)]) +
  geom_boxplot() +
  geom_jitter(width=.1, height=.1, alpha=.1) +
  ylab('# species gained/lost') +
  xlab('') +
  guides(color=F) +
  #labs(color='Transition\ntype') +
  scale_x_discrete(labels=c('not shrub\n encroached','shrub\n encroached')) +
  ggtitle('Species richness change: 1928 to 2016') +
  theme_bw()
box_richness_shrubtransition
#ggsave(box_richness_shrubtransition, filename='Figures/richness/species_richness_comparison_1928_2016.png', width=4, height=3)
```

The boxplot of the data suggests that quadrats in shrub-encroached areas lost more species. 

```{r avg loss, echo=F}
# what are the average values in 2016 and 1928?
richness_comparison %>%
  group_by(shrubland) %>%
  summarize(mean=mean(sprich_change),
            sd=sd(sprich_change))
```

T-test:

```{r ttest, echo=F}
sprich.lm = lm(sprich_change ~ shrubland, data=richness_comparison)
summary(sprich.lm)

# Augment model with the data
sprich.res = augment(sprich.lm, richness_comparison)

# Residual plot colored by shrubland
ggplot(sprich.res, aes(x = .fitted, y = .resid, col = shrubland)) +
  geom_jitter(width=.05) +
  geom_hline(yintercept = 0)
 
# Histogram of residuals
ggplot(sprich.res, aes(x = .resid)) +
  geom_histogram(fill = 'steelblue', color = 'black')
```

## Site-wide species richness
Did the average loss of 1 species per quadrat translate to site-wide species richness loss?

```{r site richness, echo=F}
# read data files and proces data
veg_selected_noblank = read.csv('../data/all_species_counts_cover.csv', stringsAsFactors = F)
dates = read.csv('../data/quadrats_dates_for_analysis.csv', stringsAsFactors = F)
splist = read.csv('../../JRN_quadrat_datapaper/Plants/Jornada_quadrat_species_list.csv', stringsAsFactors = F)

usedspecies = splist %>% dplyr::filter(species_code %in% veg_selected_noblank$species)
# just "known" species -- includes codes identified to genus (e.g. SPORO) but not total unknowns (e.g. UNKPG)
knownspecies = usedspecies %>% dplyr::filter(species !='')

# remove unknowns, shrub and subshrub species
veg_selected_noblank_noshrub = dplyr::filter(veg_selected_noblank, !form %in% c('SHRUB','S-SHR'), species %in% knownspecies$species_code)

# combine veg with dates so blank quadrats will be included
veg_selected_noshrub = veg_selected_noblank_noshrub %>%
  mutate(present = rep(1)) %>%
  merge(dates, all=T)

# how many species at site level on the 86 quadrats that were measured in 1928 and 2016
site_richness_1928 = veg_selected_noshrub %>%
  dplyr::filter(project_year==1928, quadrat %in% richness_comparison$quadrat) %>%
  dplyr::select(species, form, category) %>%
  unique() %>%
  dplyr::filter(!is.na(species))
print(paste('# species in 1928:',length(site_richness_1928$species)))
# 31 species: 12 grass and 19 forb

site_richness_2016 = veg_selected_noshrub %>%
  dplyr::filter(project_year==2016, quadrat %in% richness_comparison$quadrat) %>%
  dplyr::select(species, form, category) %>%
  unique() %>%
  dplyr::filter(!is.na(species))
print(paste('# species in 2016:',length(site_richness_2016$species)))
# 26 species: 11 grass and 15 forb

```